Ownership types in theory and practice (in Rust)

### 2022-11-29 notes.tex
Started writing paper
More writing progress
Lots of progress on rust section

@@

% - why do we need ownership types, what do they do?

@@

% source: Definition of memory safety\cite{memory1}.
% source: Why memory safery is important\cite{nsa-memory-safety}.

@@

% - when/in which language first introduced?
% - why slow adoption?

 % - TODO: source: Standard ML, C++?

% - source: \cite{understanding-evolving-rust}
% - source: \cite{ownership-types-survey}
% Notes from Ownership Types Survey\cite{ownership-types-survey}

% before: simply protect fields of an object from external access
% with OT: also protect objects stored in those fields by claiming ownership of and access to other objects
% problem OTs solve: object aliasing in OOP
% bugs due to unintentional aliases, difficult to find
% potential solutions: banning aliases, clearly advertising aliases, manage and control its effects
% page 16:
% 1998: flexible alias protection:
% limit visibility of changes to objects via aliases
% by limiting where alias could propagate
% by limiting the changes that could be observed through alias
% didn't feature "this"
% evolved into original OT:
% owners-as-dominators invariant:
% - heap divided into next regions (ownership contexts)
% - OC is set of objects
% - organised hierarchically: each owned object inside its owner, nesting relationship
% - each object has one single context
% - receives permissions to reference objects in other contexts through type

% core concept: object ownership: objects owner by other object or entities (global owners, stack-based owners)
% objects in same ownership context: siblings or peers

% control aliasing by removing aliasing
% unique references: only reference to object in system
% borrowing: 
% - reference to owned object temporarily passed to another object (for method call)
% - method over => all temporary references vanish/unusable

% control aliasing by controlling effects of aliasing
% immutability: state cannot change

% owners-as-dominators
% - program heap is tree structured, object is inside its owner
% - external access only through interface, i.e. owner
% - rules for a to validly access b: 1. a is owner of b; 2. a and b are siblings; 3. b is outside a
% - "a is inside b": "a owns b": a = b or a is owned by b, transitive
% - "a is outside b": converse of inside: "b is inside a"
% - advantages: simple, clear and strong garantee, easy to reason about properties of code
% - disadvantages: no aliasing => no common idioms; programming more difficult

% owners-as-modifiers
% - weaker form of owners-as-dominators, allows read-only references
% - read-only reference: only used to read fields, call pure methods (may not modify any existing object, including receiver)
% - aliasing unrestricted, but only owner can modify
% - usecase: requirements from verification of object-oriented programs
% - rules for a to validly reference b through reference r: either
% 1. a is owner of b
% 2. a and b are siblings
% 3. b is outside a
% 4. r is read-only reference and only pure methods can be called

% ownership domains:

% owners-as-ombudsmen:
% - problem: strong topological requirement: tree structure: some idioms don't fit in tree structure
% - allow multiple objects to define a shared aggregate owner
% - rules for a to validly reference b through reference r: either
% 1. a is owner of b
% 2. a and b are siblings
% 3. b is outside a
% 4. a is owned by aggregate owner b

% QUESTIONS:
% - difference between aliases and references
% - difference between "a is owner of b" and "b is outside a"
% - how to reference that a whole paragraph came from one source?

@@

\subsection{Implementation in Rust}
% - borrow checker

% Notes from rust-book\cite{rust-book}.
% Ownership definition: set of rules that govern how Rust program manages memory
% Compiler checks rules, if violated program doesn't compile
% Purpose of Rust ownership is to mainly manage heap data
% Ownership rules:
% - each value in Rust has an owner
% - there can only be one owner at a time
% - when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
% Definition scope:

% \begin{lstlisting}
% {                      // s is not valid here, it’s not yet declared
%     let s = "hello";   // s is valid from this point forward
%     // do stuff with s
% }                      // this scope is now over, and s is no longer valid
% \end{lstlisting}

% data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied

>>>>

% QUESTIONS:
% - difference between aliases and references
% - difference between "a is owner of b" and "b is outside a"
% - how to reference that a whole paragraph came from one source?
% - how to incorporate tables? What tables could be relevant in my paper?
% - Online article: Should I cite the actual pdf or the website that brought me to the pdf
% - british english vs american english spelling of same words?
% - should i really call it "Main part"? Or split up main part into own sections?
% - Is what i have now on owners-as-... what should go in state of the art?

@@

- why do we need ownership types, what do they do?

@@

source: Definition of memory safety\cite{memory1}.
source: Why memory safery is important\cite{nsa-memory-safety}.

@@

- when/in which language first introduced?
- why slow adoption?

 - TODO: source: Standard ML, C++?

- source: \cite{understanding-evolving-rust}
- source: \cite{ownership-types-survey}
Notes from Ownership Types Survey\cite{ownership-types-survey}

before: simply protect fields of an object from external access
with OT: also protect objects stored in those fields by claiming ownership of and access to other objects
problem OTs solve: object aliasing in OOP
bugs due to unintentional aliases, difficult to find
potential solutions: banning aliases, clearly advertising aliases, manage and control its effects
page 16:
1998: flexible alias protection:
limit visibility of changes to objects via aliases
by limiting where alias could propagate
by limiting the changes that could be observed through alias
didn't feature "this"
evolved into original OT:

owners-as-dominators invariant:
- heap divided into next regions (ownership contexts)
- OC is set of objects
- organised hierarchically: each owned object inside its owner, nesting relationship
- each object has one single context
- receives permissions to reference objects in other contexts through type

core concept: object ownership: objects owned by other object or entities (global owners, stack-based owners)
objects in same ownership context: siblings or peers

control aliasing by removing aliasing
unique references: only reference to object in system
borrowing: 
- reference to owned object temporarily passed to another object (for method call)
- method over => all temporary references vanish/unusable

control aliasing by controlling effects of aliasing
immutability: state cannot change

owners-as-dominators
- program heap is tree structured, object is inside its owner
- external access only through interface, i.e. owner
- rules for a to validly access b: 1. a is owner of b; 2. a and b are siblings; 3. b is outside a
- "a is inside b": "a owned by b": a = b or a is owned by b, transitive
- "a is outside b": converse of inside: "b is inside a"
- advantages: simple, clear and strong garantee, easy to reason about properties of code
- disadvantages: no aliasing => no common idioms; programming more difficult

owners-as-modifiers
- weaker form of owners-as-dominators, allows read-only references
- read-only reference: only used to read fields, call pure methods (may not modify any existing object, including receiver)
- aliasing unrestricted, but only owner can modify
- usecase: requirements from verification of object-oriented programs
- rules for a to validly reference b through reference r: either
1. a is owner of b
2. a and b are siblings
3. b is outside a
4. r is read-only reference and only pure methods can be called

ownership domains:

owners-as-ombudsmen:
- problem: strong topological requirement: tree structure: some idioms don't fit in tree structure
- allow multiple objects to define a shared aggregate owner
- rules for a to validly reference b through reference r: either
1. a is owner of b
2. a and b are siblings
3. b is outside a
4. a is owned by aggregate owner b

@@

\subsection{Implementation in Rust}
- borrow checker

Notes from rust-book\cite{rust-book}.
Ownership definition: set of rules that govern how Rust program manages memory
Compiler checks rules, if violated program doesn't compile
Purpose of Rust ownership is to mainly manage heap data
TODO: contradiction to below: value can be invalid in scope
Ownership rules:
- each value in Rust has an owner
- there can only be one owner at a time
- when the owner goes out of scope, the value will be dropped (value is valid in scope, invalid out of scope)
Definition scope:

\begin{lstlisting}
{                      // s is not valid here, it’s not yet declared
    let s = "hello";   // s is valid from this point forward
    // do stuff with s
}                      // this scope is now over, and s is no longer valid
\end{lstlisting}

data types of known size at compile time (int, float, bool, char): stored on stack, popped off stack when out of scope, quickly and trivially copied
heap: mutable data
stack: immutable data, size known at compile time
requirements of memory allocation system:
1. memory must be requested from memory allocator at runtime
2. need way of returning/freeing memory when done using it

First part is trivial
Second part: other solutions:
1. garbage collector: keeps track of and cleans up memory isn't used anymore (we don't need to think about it/do anything)
2. without GC, traditional: explicitly free memory
- difficult
- if forget, waste memory
- if too early, invalid variable
- if free twice, bug
=> need to pair one allocate with exactly one free
Rust: memory is automatically freed once variable that owns it goes out of scope
Rust calls `drop` automatically at end of every scope

Move:
simple (find different word) data that is stored on stack: data is copied
example:
\begin{lstlisting}
let x: i32 = 42;
let y: i32 = x;
\end{lstlisting}

For datastructures stored on the heap like String:

\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
\end{lstlisting}

similar to shallow copy of fields stored on stack (pointer to memory in heap, maybe length or capacity as well).
Assumption: both x and y point to the same memory in the heap.
what happens when they go out of scope, what is freed?
One object cannot have two owners.
In rust: variable x is moved into y = y becomes the new owner of the object and x is invalid

### 2022-11-29 paper.tex
Lots of progress on rust section

% TODO: second source to provide more detail of each owners-as-...
% TODO: find more varied words to classify these approaches as: principle, approach, model

@@

Another kind is the owners-as-modifiers model, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
In essence, aliasing is unrestricted, but only the owner of an object can modify it.
% TODO: unclear sentence structure
Owners-as-modifiers extend the rules specified for owners-as-dominators.
If none of the rules that were specified for owners-as-dominators are adhered to, for \emph{a} to validly reference \emph{b} through reference \emph{r}, \emph{r} has to be a read-only reference and only pure methods can be called on it.
% TODO: more detail

@@

% TODO: definition of ombudsmen for context
A last flavour of ownership types is the owners-as-ombudsmen approach, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
% TODO: examples
This is achieved by allowing multiple objects to define a shared aggregate owner.
% TODO: what is aggregate owner
In addition to the rules stated for owners-as-dominators, \emph{a} has to be owned by aggregate owner \emph{b} for \emph{a} to validly reference \emph{b} through reference \emph{r}.

@@

>>>>

% TODO: second source to provide more detail of each owners-as-...
% TODO: find more varied words to classify these approaches as: principle, approach, model
% TODO: maybe add numbered list (like in source material) for the rules like in the Rust section

@@

Another kind is the owners-as-modifiers model, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
In essence, aliasing is unrestricted, but only the owner of an object can modify it.
% TODO: unclear sentence structure
Owners-as-modifiers extend the rules specified for owners-as-dominators.
For \emph{a} to validly reference \emph{b} through reference \emph{r}, onwers-as-modifiers follows the same rules as owners-as-dominators.
The only addition is that, if none of the owners-as-dominators rules are upheld, \emph{r} can also be valid by being a read-only reference and only pure methods being called on it.
% TODO: more detail

@@

% TODO: definition of ombudsmen for context
A last flavour of ownership types is the owners-as-ombudsmen approach, which allows idioms that do not adhere to the strong topological structure of a tree structure to be implemented.
% TODO: examples
This is achieved by allowing multiple objects to define a shared aggregate owner.
% TODO: what is aggregate owner
Extending the rules stated for owners-as-dominators, \emph{a} can also validly reference \emph{b} if \emph{a} is owned by aggregate owner \emph{b}.

@@

There are two major requirements that a memory allocation system must fulfill:
At runtime, memory can be requested from the memory allocator and there exists a mechanism for memory to be freed by the memory allocator.

The first requirement is trivial.

% TODO: find source for both garbage collection and manual freeing
One possible implementation of the second requirement is using a garbage collector, which keeps track of and cleans up memory that is no longer used.
% TODO: find source for GC disadvantages
This, however, has a runtime performance overhead and can lead to a non-deterministic cleanup of resources.

% TODO: find source for manual free disadvantages
Another more traditional solution is requiring memory to be freed explicitly.
This is a notorious source of bugs.
If memory is never freed because it was forgotten, memory is wasted.
If it is freed multiple times or accessed again after having been freed, it will result in undefined behaviour.

The ideal solution pairs one \emph{allocate} with exactly one \emph{free}.
Rust uses neither a garbage collector nor requires memory to be freed manually.
Instead, the concept of ownership is leveraged to determine when to free memory.

The Rust book \cite{rust-book} defines ownership as a set of rules that govern how it manages memory.
The compiler checks whether these rules are adhered to, and does not compile the program if they are violated.
The ownership rules in Rust are as follows:
\begin{enumerate}
  \item Each value in Rust has an owner.
  \item There can only be one owner at a time.
  \item When the owner goes out of scope, the value will be \emph{dropped}.
\end{enumerate}
% Rust automatically calls the function \emph{drop} when a variable goes out of scope, which frees the memory of the value that is owned by that variable.
% TODO: is a variable really an owner of a value? Possibly change term variable
The scope of a value is the range within a program in which it is valid.
Rust automatically calls the function \emph{drop}, which frees the associated memory, when a variable goes out of scope.

The following describes the concepts of how variables and memory interact in Rust.
\cite{rust-book}

\subsubsection{Move}

% TODO: wording of value and types used interchangedly here, maybe incorrect
First, one must differentiate between values that are stored on the stack and those stored on the heap.

The primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack since their sizes are small and known at compile time.
Values are stored on the heap if their size is dynamic and therefore not known at compile time.

The different representations of strings in Rust provide a good example.
Rust supports the immutable string literals \emph{\&str}, which are stored on the stack, and the type \emph{String}, which is stored as a vector of bytes, heap allocated, and therefore growable. \cite{rust-by-example}

The \emph{String} type in Rust is a good example of such a type.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
\end{lstlisting}

% This, however, applies only to data types that are stored on the heap.

In the following piece of Rust code, one variable is assigned to another.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
\end{lstlisting}
Since their sizes are known at compile time, they are simply copied into a new variable instead of moved.

\subsubsection{Clone}

By default and by design, Rust will never automatically create deep copies of data stored on the heap.
To do so, one must explicitly call the \emph{clone} method on an object.

\cite{rust-book}
